Composition-climate paper

### 2015-10-27 Composition_Ecology.Rmd
More references to build up the introduction.  Mostly trying to clarify the compounding/confounding and contradicting concepts.  Added some climate references too.
Clarified the hypotheses examined in the paper.

The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].  Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.  By deliniating three tpyes of interaction, compunding, confounding and contradicting, it is possible to understand both the nature of the interaction and the potential effect on species resiliance and response to broader regional change.

Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.  In the case of warming a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.  Confounding effects may cause little overall effect on species distributions since in this case the land use pressure falls within the central distribution for the environmental variable, they may increase the variability of predictive models, but do little to the overall envelope.  Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in the warmer tail of the distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the cooler end of the tail excludes the taxa from those regions.

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al. [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical]. To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies], we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

Interaction between environmental gradients and land use patterns vary by region and environmental variable.  We might expect that species in the north experience a confounding land use effect with respect to temperature, but, with respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect this effect to be compounding.  For southern tree species land use conversion is primarily, and has primarily occured along the southern margins of the region, parallel to temperature and precipitation gradients.  This should result in contradicting land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the twin effects of land use change and regional climate change.

Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting land use effects may result in the impression of narrower climatic tolerances.  For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envolope may result in underprediction of species resiliance to climate change.

We use gridded climate data products and estimates of pre-settlement vegetation to develop climate-vegetation relationships for 15 major tree taxa in the upper Midwestern United States to understand how climate and land use change have interacted since Euro-American settlement to affect species distributions in the region, and to examine the possible implications of these shifts to future estimates of species resillience to climate change across the region.

@@

Another complication is the structure of the FIA dataset.  While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].  As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA. This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations. [needs refs].

>>>>

The interaction of land use change and climatic change has been of interest to conservation managers and scientists [e.g., @pyke2004habitat] since managed natural areas often represent a biased selection from the available environmental space [@scott2001nature].  Pyke [-@pyke2004habitat] lays out a useful framework for assessing the interaction of climate and land use change.  By deliniating three tpyes of interaction: (1) compounding, (2) confounding and (3) contradicting, it is possible to understand both the nature of the climate/land use interaction and the potential effect on apparent species resiliance and future response to broader regional change.

Compounding effects occur when both climatic change and land use change are biased in the same direction, *e.g.*, warming climate is compounded by land use change that is biased towards warmer environments.  In the case of warming, a species may lose climatic suitability along the warmest margins of its range at the same time as land use pressure increases, hastening decline in response to warming.  This pattern is seen for butterfly species in the northern Sierra Nevada mountains of California, where warming is compounded by land use conversion at the lowest elevations [@forister2010compounded].  Confounding effects may cause little overall effect on species distributions since land use pressure falls within the central distribution for the environmental variable. Confounding effects may increase the variability of predictive models, but do little to the overall envelope.  Garry oak distributions in the Pacific Northwest may suffer from confounding effects.  The spatial extent of the species has likely changed little, but it has been excluded from the deep soil sites it formerly occupied as a result of land use conversion for urban development and agriculture [@macdougall2004defining]. Contradicting effects result in a narrower climate niche for taxa, where warming forces the distribution mean higher as populations in one tail of the climate distribution experience greater environmental pressure, and, secondarily environmentally biased land use pressure in the other end of the tail excludes the taxa from those regions.  Here we might think of downward migrations of some bird species in the Sierra Nevadas in response to shifts in precipitation [@tingley2012push], where downward migration may be limited by land use conversion in the lowlands.

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al. [-@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical]. To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies], we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

Interaction between environmental gradients, land use patterns and taxa are likely to vary by region and environmental variable of interest.  Species in the north may experience a confounding land use effect with respect to temperature, northward shifts due to warming temperatures may not have significantly manifested themselves, and land use conversion is likely to be heterogenous with respect to temperature since there is little agriculture in northern Wisconsin and northeastern Minnesota.  Thus, with respect to temperature we expect to see no shift in distribution along temperature gradients.

With respect to precipitation, particularly toward the forest-prairie margin, where land use conversion continues [@wright2013recent] we would expect to see compounding effects of land use and climate change. Northern species should be shifting away from drier habitat while land use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient.

For southern tree species, land use conversion is primarily, and has primarily occured along the southern margins of the region, parallel to temperature and precipitation gradients. This should result in compounding land use effects for all taxa with distributions in the region, resulting in narrower species distributions due to the loss of suitability and land use conversion along the southern margin of the species' range.

Critically, given the importance of modern distributional datasets for predicting species range shifts, contradicting and confounding land use effects may result in the impression of narrower climatic tolerances, or greater predictive uncertainty. Tree species are clearly responding to climate change within forested landscapes, showing, for example, differences in the distributions of mature trees and seedlings [@monleon2015evidence], but modern correlational analysis leaves out historical distributions that can further improve our ability to predict future distributions and management implications of species shifts. For species that reach their northern or western range limits in the upper Midwestern United States (e.g., hemlock, beech, &tc.) this may be critical, since a narrowing of the apparent environmental envelope, dut to the twin action of land use change and changing climate, may result in underprediction of species resiliance to climate change.

We use gridded climate data products (PRISM) and estimates of pre-settlement and modern forest vegetation [@goring2015preset] to develop climate-vegetation relationships for 15 major tree taxa in the upper Midwestern United States. By examining the interactions between climate, species distributions and land use change since EuroAmerican settlement we examine the possible implications of range shifts for future estimates of species resillience to climate change across the region.

@@

Another complication is the structure of the FIA dataset.  While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].  As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA. This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations, agency that ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].
